A Comparative Quality Metric for Untargeted Fuzzing with Logic State Coverage

### 2024-06-02 paper/main.tex
sec2

%\pp{Normality of program behaviors}
%
Program behaviors can be divided into two classes depending on their
\emph{normality}: normal and abnormal. Normal program behaviors are determined
by a regular control flow and terminate at program exit points.  Abnormal
program behaviors, on the other hand, are terminated unexpectedly by
exceptions.\footnote{Sanitization can be seen as bringing (potentially) silent
\emph{unintended} behaviors (e.g., integer-overflow or use-after-free) to the
\emph{abnormal} territory.}

\pp{Special case: single-thread programs}
%
For a single-thread program where every basic block always executes the
contained instructions atomically, program behaviors are exclusively
determined by the history of two kinds of conditions: branch and exceptional.
Branch conditions (or \emph{branches}) decide the regular control flow between
basic blocks. Exceptional conditions (or \emph{exceptions}), on the other hand,
interrupt the control flow and abnormally terminate the
program.\footnote{C++-style exceptions, despite its name, can be regarded as a
form of branch conditions as they transfer the control flow to another basic
block.} Notice that exceptions may be implicit and not specified in the
program (e.g., $\mathrm{divisor}=0$ in divide-by-zero).

%Notice that exceptions may be implicit, unlike branches that are always explicit
%in the program; exceptions such as assertions are explicitly specified in
%the program, but others such as divide-by-zero (i.e., $\mathrm{divisor}=0$) or
%buffer-overflow (i.e., $\mathrm{offset} \ge \mathrm{size}$) are implicit and
%only enforced by the system.

Formally, let $\mathcal{T}_b$ and $\mathcal{T}_e$ be the history of branches and
exceptions. Then, for a single-thread program, there exists a function
$f(\mathcal{T}_b, \mathcal{T}_e) = h$ that one-to-one maps a pair
$(\mathcal{T}_b, \mathcal{T}_e)$ to a certain program behavior $h$.
%
%\begin{equation}
%  f(\mathcal{T}_b, \mathcal{T}_e) = h \iff f: \mathbb{T}_b \times \mathbb{T}_e \rightarrow \mathbb{H}.
%\end{equation}

@@

\pp{Completeness of untargeted fuzzing}
%
As the procedure of untargeted fuzzing can be seen as maximizing observed
interesting behaviors in the program behavior space, the portion of such 
behaviors in the space directly indicates the \emph{completeness} of the
procedure: the more it checks such behaviors, the less the unknown
abnormal behaviors are expected to be left.

Based on this, we can define the \emph{relative completeness} between
different untargeted fuzzing techniques as follows: given the same amount of
time and computing resource, \textbf{how many more interesting program behaviors
has it observed than others?} We argue that relative completeness is a
\emph{strong quality metric} for untargeted fuzzing as it closely indicates how
much more a fuzzer reduces the expectation of unknown abnormal behaviors than
others.

>>>>

\pp{Program conditions}
%
Program conditions are the instructions that affect the control flow and
transform the program behavior accordingly. They are further
broken down to two kinds: branch and exceptional.
Branch conditions (or \emph{branches}) decide the regular control
flow between basic blocks. Exceptional conditions (or \emph{exceptions}), on the
other hand, interrupt the control flow and abnormally terminate the
program.\footnote{C++-style exceptions, despite its name, can be regarded as a
form of branch conditions as they transfer the control flow to another basic
block.} 
%Notice that exceptions may be implicit and not specified in the program
%(e.g., $\mathrm{divisor}=0$ in divide-by-zero).

\pp{Normality of program behaviors}
%
Program behaviors can be divided into two classes depending on the satisfied
program conditions: normal and abnormal. Normal program behaviors only satisfy
regular branches and terminate at program exit points. Abnormal
program behaviors, on the other hand, satisfy at least one exception and
terminate at non-exit points. Sanitization
\cite{yun2016apisan,serebryany2012asan,kasan,ktsan,kubsan,stepanov2015msan,ubsan} can be seen as bringing
\emph{unintended} behaviors (e.g., integer-overflow or use-after-free) to the
\emph{abnormal} territory by adding exceptions.

\pp{Uniqueness of program behaviors}
%
For a single-thread program, 
%where every basic block always executes the contained instructions atomically, 
program behaviors are exclusively determined by the history of program
conditions. This is because a single-thread program always executes the 
instructions sequentially until the next program condition, making every
portion of the instruction stream \emph{implied} by terminating program
conditions.

Formally, let $\mathcal{T}_b$ and $\mathcal{T}_e$ be the history of branches and
exceptions. Then, for a single-thread program, there exists a function
$f(\mathcal{T}_b, \mathcal{T}_e) = h$ that uniquely maps a pair
$(\mathcal{T}_b, \mathcal{T}_e)$ to a certain program behavior $h$.
%
%\begin{equation}
%  f(\mathcal{T}_b, \mathcal{T}_e) = h \iff f: \mathbb{T}_b \times \mathbb{T}_e \rightarrow \mathbb{H}.
%\end{equation}

@@

\pp{Quality of untargeted fuzzing}
%
As the procedure of untargeted fuzzing can be seen as maximizing observed
interesting behaviors in the program behavior space, the portion of such 
behaviors in the space directly indicates the \emph{quality} of the
procedure: the more it checks such behaviors, the less the unknown
abnormal behaviors are expected to be left.

Based on this, we can define the \emph{comparative quality} between
different untargeted fuzzing techniques as follows: given the same amount of
time and computing resource, \textbf{how many more interesting program behaviors
can it observe than others?} We argue that this is a \emph{strong quality
metric} for untargeted fuzzing as it directly compares the reduced expectation
of unknown abnormal behaviors.
